---
title: "Parse blast output"
author: "Andreas Evenstad"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load packages
```{r}
library(tidyverse)
library(ComplexUpset)
getwd()
```

# Load BLAST output
```{r}
blast_output <- read.table("methanotrophs_all_hits.blastout", header = FALSE, sep = "\t") %>%
  distinct()
colnames(blast_output) = c("query", "subject", "identity", "alignment_length", "mismatches", "gap_opens", "q_start", "q_end", "s_start", "s_end", "evalue", "bit_score")
```

## Clean up subject names
```{r}
# Clean up species names
blast_output <- blast_output %>%
  mutate(subject = gsub("M.aurea_v1_16s_rRNA_1", "Methylocapsa_aurea", subject)) %>%
  mutate(subject = gsub("Methylocapsa_gorgona_MG08_T", "Methylocapsa_gorgona", subject)) %>%
  mutate(subject = gsub("Methylocapsa_palsarum_NE2_T_\\(KP715289\\)", "Methylocapsa_palsarum", subject)) %>%
  mutate(subject = gsub("MROSv2_16s_rRNA_1", "Methylocystis_rosea", subject)) %>%
  mutate(subject = gsub("M.acidiphila_16s_rRNA_1", "Methylocapsa_acidiphila", subject)) %>%
  mutate(subject = gsub("M.acidiphila_16s_rRNA_2", "Methylocapsa_acidiphila", subject)) %>%
  mutate(subject = gsub("Methylocella_tundrae_T4", "Methylocella_tundrae", subject)) %>%
  mutate(subject = gsub("Methylocystis_echinoides_IMET", "Methylocystis_echinoides", subject)) %>%
  mutate(subject = gsub("Methylocystis_parvus_OBBP", "Methylocystis_parvus", subject)) %>%
  mutate(subject = gsub("Methyloferula_stellata_AR4", "Methyloferula_stellata", subject)) %>%
  mutate(subject = gsub("Methylosinus_sporium_AB02", "Methylosinus_sporium", subject)) %>%
  mutate(subject = gsub("M.trich_v2_16s_rRNA_1", "Methylosinus_trichosporium", subject))
```



# Get metadata
## Load SRA metadata
```{r load_biosample_metadata, echo=FALSE}
biosample_metadata <- read.table("../01_ncbi_search/summary_table_final.tsv", 
                header = TRUE, sep = "\t", quote="", stringsAsFactors=FALSE, fill = TRUE)
```


## Load SRA run and SRA sample ids
```{r load_sra_run_sample_ids, echo=FALSE}
sra_run_sample_ids <- read.table("../02_fetchngs/SRA_run_and_sample_accession.txt", sep = ",", header = TRUE)
```


## Merge the metadata with the SRA run and sample ids
```{r merge_metadata, echo=FALSE}
metadata <- biosample_metadata %>%
  left_join(sra_run_sample_ids, by = c("SRA" = "secondary_sample_accession"))
```

## Add metadata to blast results
```{r add_metadata, echo=FALSE}
blast_output_metadata_df = blast_output %>%
  mutate(sra_id = gsub("\\..*", "", query)) %>%
  left_join(metadata, by = c("sra_id" = "run_accession")) %>%
  filter(!is.na(BioSample)) %>%
  distinct()
```


# Process blast results
## Extract unique hits
```{r}
# Filter the data frame to only include rows where the query name appears once
unique_query_target <- blast_output_metadata_df %>%
  add_count(query) %>%    # Add a column 'n' with the count of each query
  filter(n == 1) %>%      # Keep only rows where 'n' (count) is 1
  select(-n)            # Optionally remove the count column if not needed


# Identify queries where the subject is either M.acidiphila_16s_rRNA_1 or M.trich_v2_16s_rRNA_1
filter_out_queries = unique_query_target %>%
  filter(subject %in% c("Methylocapsa_acidiphila", "Methylosinus_trichosporium")) %>%
  pull(query)

# Remove species not of interest and retain the best hit for each query
blast_output_unique_df = unique_query_target %>%
  filter(!query %in% filter_out_queries) %>%
  mutate(sra_id = gsub("\\..*", "", query))


# Print number of unique queries and hits
cat("Number of unique queries (only one hit):", length(unique(blast_output_unique_df$query)), "\n")
cat("Number of rows (hits) for unique queries:", nrow(blast_output_unique_df), "\n")

# Count the number of reads and SRAs per species
unique_species_counts = blast_output_unique_df %>%
  group_by(subject) %>%
  summarise(
    reads = n(),
    SRAs = n_distinct(gsub("\\..*", "", query))
  )

cat("Unique hits - Number of reads and SRAs per species:\n")
print(unique_species_counts)


# Keep the best hit for each SRA ID-subject pair
blast_output_unique_df = blast_output_unique_df %>%
  group_by(sra_id, subject) %>%
  filter(bit_score == max(bit_score)) %>%
  distinct(sra_id, subject, .keep_all = TRUE)

```

## Save unique hits table
```{r}
write.table(blast_output_unique_df, "unique_hits_metadata.tsv", sep = "\t", quote = TRUE, row.names = FALSE, col.names = TRUE, na = "NA")
```

## Multiple hits
```{r}
# Multiple hits
cat("\nProcessing multiple hits...\n")
multiple_hits_queries = names(which(table(blast_output_metadata_df$query) > 1))
blast_output_multiple = blast_output_metadata_df[blast_output_metadata_df$query %in% multiple_hits_queries, ] %>%
  distinct()

# Print number of queries with multiple hits
cat("Number of queries with multiple hits:", length(multiple_hits_queries), "\n")
cat("Number of rows (hits) for queries with multiple hits:", nrow(blast_output_multiple), "\n")

# identify queries where subject is either M.acidiphila_16s_rRNA_1 or M.trich_v2_16s_rRNA_1
filter_out_queries_multi = blast_output_multiple %>%
  filter(subject == "Methylocapsa_acidiphila" | subject == "Methylosinus_trichosporium") %>%
  pull(query)

cat("Number of queries filtered out due to outgroup species:", length(filter_out_queries_multi), "\n")

# Remove queries where subject is either M.acidiphila_16s_rRNA_1 or M.trich_v2_16s_rRNA_1
blast_output_multiple_hits_df = blast_output_multiple %>%
  filter(!query %in% filter_out_queries_multi) %>%
  group_by(query, subject) %>%
  filter(bit_score == max(bit_score)) %>%
  distinct(query, subject, .keep_all = TRUE) %>%
  mutate(sra_id = gsub("\\..*", "", query))

# Print number of hits remaining after filtering out outgroup species and retaining best hits
cat("Number of hits after filtering out outgroup species and retaining best hits:", nrow(blast_output_multiple_hits_df), "\n")

# Count the number of reads and SRAs per species for multiple hits
multiple_species_counts = blast_output_multiple_hits_df %>%
  group_by(subject) %>%
  summarise(
    reads = n(),
    SRAs = n_distinct(sra_id)
  )

cat("Multiple hits - Number of reads and SRAs per species:\n")
print(multiple_species_counts)
```


## Save multiple hits table
```{r}
write.table(blast_output_multiple_hits_df, "multiple_hits_metadata.tsv", sep = "\t", quote = TRUE, row.names = FALSE, col.names = TRUE, na = "NA")
```

## Save all hits table
```{r}
all_hits_df <- rbind(blast_output_unique_df, blast_output_multiple_hits_df)

write.table(all_hits_df, "all_hits_metadata.tsv", sep = "\t", quote = TRUE, row.names = FALSE, col.names = TRUE, na = "NA")
```


## Create upset plot for multiple hits matching M.acidiphila and M.trichisprium
```{r}
# Prepare data for upset plot
multiple_hits_upset = blast_output_multiple %>%
  filter(query %in% filter_out_queries_multi) %>%
  select(query, subject) %>%
  group_by(query, subject) %>%
  mutate(value = 1) %>%
  distinct() %>%
  pivot_wider(names_from = subject, values_from = value, values_fill = 0) %>%
  column_to_rownames(var = "query")



# Create the upset plot
png(file = "final_analysis/04_blast_16S_methanotrophs/upset_plot_outgroup_hits.png",   # The directory you want to save the file in
    #onefile=F, # If TRUE, all plots are put in one file. If FALSE, a new file is created for each plot.
    width = 3840, # The width of the plot in inches
    height = 2160,
    res = 300) # The height of the plot in inches

ComplexUpset::upset(
  multiple_hits_upset,
  intersect = colnames(multiple_hits_upset),
  min_size = 1,            # Minimum size for displayed intersections
  n_intersections = 5     # Limit to top 5 intersections
)

dev.off()
```

```{r}
# Prepare data for upset plot
multiple_hits_upset = blast_output_multiple %>%
  filter(!query %in% filter_out_queries_multi) %>%
  select(query, subject) %>%
  group_by(query, subject) %>%
  mutate(value = 1) %>%
  distinct() %>%
  pivot_wider(names_from = subject, values_from = value, values_fill = 0) %>%
  column_to_rownames(var = "query")

# Create the upset plot
png(file = "upset_plot_for_multiple_hits.png",   # The directory you want to save the file in
    #onefile=F, # If TRUE, all plots are put in one file. If FALSE, a new file is created for each plot.
    width = 3840, # The width of the plot in inches
    height = 2160,
    res = 300) # The height of the plot in inches

ComplexUpset::upset(
  multiple_hits_upset,
  intersect = colnames(multiple_hits_upset),
  min_size = 1,            # Minimum size for displayed intersections
  n_intersections = 5     # Limit to top 5 intersections
)

dev.off()
```












